\documentclass[style=sailor,size=12pt,mode=present]{powerdot}
\usepackage{epic,array,ecltree,url,calrsfs}
\usepackage[nointegrals]{wasysym}
\usepackage{mlextra}
\usepackage{listings}
\graphicspath{ {../art/} }

\newcommand{\id}[1]{\mbox{\it #1\/}}
\newcommand{\rid}[1]{\mbox{\rm #1}}
\newcommand{\sid}[1]{\mbox{\sf #1}}
\newcommand{\bid}[1]{\mbox{\bf #1}}
\newcommand{\tinysz}[1]{\mbox{\tiny $#1$}}

\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsxtra}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{color}
\usepackage{array}
\usepackage{graphicx}
\graphicspath{ {../art/} }
\usepackage{bm}
\usepackage{tikz}
\usepackage{multicol}
\usepackage{enumitem}

\renewcommand\qedsymbol{$\blacksquare$}

%Here I define some theorem styles and shortcut commands for symbols I use often
\theoremstyle{definition}

\newtheorem{innerdefn}{Definition}
\newenvironment{defn}[1]
  {\renewcommand\theinnerdefn{#1}\innerdefn}
  {\endinnerdefn}

\newtheorem{innerexample}{Example}
\newenvironment{ex}[1]
  {\renewcommand\theinnerexample{#1}\innerexample}
  {\endinnerexample}

\newtheorem{innerthm}{Theorem}
\newenvironment{thm}[1]
  {\renewcommand\theinnerthm{#1}\innerthm}
  {\endinnerthm}

\newtheorem{innerlem}{Lemma}
\newenvironment{lem}[1]
  {\renewcommand\theinnerlem{#1}\innerlem}
  {\endinnerlem}




%\newtheorem*{defn}{Definition}
%\newtheorem{thm}{Theorem}
%\newtheorem{cor}{Corollary}
%\newtheorem*{rmk}{Remark}
%\newtheorem{lem}{Lemma}
%\newtheorem*{joke}{Joke}
%\newtheorem{ex}{Example}
%\newtheorem*{soln}{Solution}
%\newtheorem{prop}{Proposition}


\newcommand{\lra}{\longrightarrow}
\newcommand{\ra}{\rightarrow}
\newcommand{\surj}{\twoheadrightarrow}
\newcommand{\graph}{\mathrm{graph}}
\newcommand{\bb}[1]{\mathbb{#1}}
\newcommand{\Ell}{\mathscr{L}}
\newcommand{\Z}{\bb{Z}}
\newcommand{\Q}{\bb{Q}}
\newcommand{\R}{\bb{R}}
\newcommand{\C}{\bb{C}}
\newcommand{\N}{\bb{N}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\m}{\mathbf{m}}
\newcommand{\MM}{\mathscr{M}}
\newcommand{\HH}{\mathscr{H}}
\newcommand{\Om}{\Omega}
\newcommand{\Ho}{\in\HH(\Om)}
\newcommand{\bd}{\partial}
\newcommand{\del}{\partial}
\newcommand{\bardel}{\overline\partial}
\newcommand{\textdf}[1]{\textbf{\textsf{#1}}\index{#1}}
\newcommand{\img}{\mathrm{img}}
\newcommand{\ip}[2]{\left\langle{#1},{#2}\right\rangle}
\newcommand{\inter}[1]{\mathrm{int}{#1}}
\newcommand{\exter}[1]{\mathrm{ext}{#1}}
\newcommand{\cl}[1]{\mathrm{cl}{#1}}
\newcommand{\ds}{\displaystyle}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cnt}{\mathrm{ct}}
\newcommand{\osc}{\mathrm{osc}}
\newcommand{\LL}{\mathbf{L}}
\newcommand{\UU}{\mathbf{U}}
\newcommand{\support}{\mathrm{support}}
\newcommand{\AND}{\;\wedge\;}
\newcommand{\OR}{\;\vee\;}
\newcommand{\Oset}{\varnothing}
\newcommand{\st}{\ni}
\newcommand{\wh}{\widehat}
\newcommand{\mli}[1]{\mathit{#1}}
\newcommand{\ndiv}{\hspace{-3pt}\not|\hspace{2pt}}


\pdsetup{method=normal}

\title{First Order Logic: Computational Complexity and Decidable Subclasses}
\author{Foundations of Computer Science}
\date{\today}

\begin{document}
\maketitle
\section[slide=false]{Computational Complexity of FOL}
\begin{slide}[bm=,toc=]{Validity in FOL is Undecidable}
There is no decision procedure for validity in first order logic.
\begin{itemize}
\item First proved by Alonzo Church in 1936.
\item Proof uses a reduction \emph{from} the halting problem.
\item That is $\id{HALT} \leq \id{FOL-VAL}$
\begin{itemize}
\item The halting problem is proven to be undecidable.
\item Therefore, $\id{FOL-VAL}$ is undecidable.
\end{itemize}
\end{itemize}
How the reduction works:
\begin{itemize}
\item Begins with an arbitrary Turing machine $T$.
\item Constructs a formula $S_T$ in first-order logic, such that
\begin{itemize}
\item $S_T$ is valid iff $T$ halts on a blank tape.
\end{itemize}
\item This shows that a decision procedure for $\id{FOL-VAL}$ would also solve
the halting problem.
\end{itemize}
\end{slide}

\begin{wideslide}[bm=,toc=]{Church's Theorem}
\begin{thm}{12.3}[Church]

\emph{Validity in first-order logic is undecidable.}
\end{thm}
Holds even under the following restrictions:
\begin{enumerate}
\item The formulas contain only binary predicate symbols, one constant and one
unary function symbol.
\begin{itemize}
\item Ex: $\forall x (p(x,f(x)) \imp a$.
\end{itemize}
\item The formulas are logic programs.
\begin{itemize}
\item Restricted set of clauses structured to perform computations.
\item See Ben Ari chapter 11.
\end{itemize}
\item The formulas are pure (see below).
\end{enumerate}
\vspace{-2ex}
\begin{defn}{12.4}[Ben Ari]
A formula of first-order logic is \emph{pure} if it contains no function symbols
(including constants which are 0-ary functions symbols).
\end{defn}
{\bf Note:} Validity is undecidable \emph{with} or \emph{without} function symbols.
\end{wideslide}

\begin{wideslide}[bm=,toc=]{Additional Implications of Church's Theorem}
{\bf Recall:}\\
A formula is valid if and only if its complement (negation) is unsatisfiable.
\begin{itemize}
\item There is a simple reduction from a formula to its complement (negation).
\item Therefore, $\id{FOL-VAL} \leq \id{FOL-UNSAT}$ 
\item $\id{FOL-VAL}$ is undecidable.
\item Therefore, $\id{FOL-UNSAT}$ is undecidable.
\end{itemize}
\end{wideslide}

\begin{wideslide}[bm=,toc=]{Semi-decision Procedures for FOL-UNSAT}
\begin{itemize}
\item Resolution for FOL is semi-decision procedure (J.A Robinson) for FOL-UNSAT.
\item Another technique, constructing a \emph{semantic tableau} is also
semi-decision procedure.
\item Thus, $\id{FOL-UNSAT}$ is semi-decidable.
\item Since $\id{FOL-VAL} \leq \id{FOL-UNSAT}$ 
\item $\id{FOL-VAL}$ is semi-decidable. 
\end{itemize}
{\bf Recall:} 
\begin{itemize}
\item $\id{FOL-SAT}$ = $\id{FOL}$ - $\id{FOL-UNSAT}$
\item Properties of semi-decision procedures:
\begin{itemize}
\item Return ``yes'' if an element is in the target set.
\item Return ``no'' or runs forever if not.
\end{itemize}
\item Therefore, $\id{FOL-SAT}$ is \emph{neither} decidable \emph{nor}
semi-decidable.
\end{itemize}

\end{wideslide}

\begin{wideslide}[bm=,toc=]{Satisfiability vs Validity in PL and FOL}
In propositional logic:
\begin{itemize}
\item $\id{SAT}$ is NP-complete $\imp \id{UNSAT}$ is CoNP-complete. 
\item $\id{UNSAT} \leq_p \id{VAL}$, so $\id{VAL}$ is coNP-Hard.
\item $\id{INVAL}$ is in NP $\imp \id{VAL}$ is in CoNP.
\item $\id{VAL}$ is in CoNP and $\id{VAL}$ is CoNP-Hard $\imp \id{VAL}$ is Co-NP
complete.
\item Therefore, in PL, $\id{VAL}$ is harder than $\id{SAT}$ unless $NP = CoNP$.
\end{itemize}
In first-order logic:
\begin{itemize}
\item $\id{VAL} \leq_p \id{UNSAT}$
\begin{itemize}
\item $\id{VAL}$ is undecidable $\imp \id{UNSAT}$ is undecidable. 
\item $\id{UNSAT}$ is semi-decidable $\imp \id{VAL}$ is semi-decidable. 
\end{itemize}
\item $\id{UNSAT}$ is semi-decidable $\imp \id{SAT}$ \emph{not} semi-decidable. 
\item Therefore, in FOL, $\id{SAT}$ is harder than $\id{VAL}$.
\end{itemize}
\end{wideslide}


\section[slide=false]{Subclasses of First-Order Logic}
\begin{slide}[bm=,toc=]{Decidable Subclasses}
\begin{thm}{12.5}
There are decision procedures for the validity of pure PCNF formulas whose
prefixes are of one of the following forms (where $m,n \geq 0$):
\[ \forall x_1\cdots \forall x_n \exists y_1 \cdots \exists y_m, \]
\[ \forall x_1\cdots \forall x_n \exists y \forall z_1 \cdots \forall z_m, \]
\[ \forall x_1\cdots \forall x_n \exists y_1 \exists y_2 \forall z_1 \cdots \forall z_m. \]
These classes are conveniently abbreviated: 

\[    \forall^*\exists^*, \forall^*\exists \forall^*,   \forall^*\exists \exists \forall^*\]
\end{thm}
\end{slide}

\begin{wideslide}[bm=,toc=]{Undecidable Subclasses}
\begin{thm}{12.6}
There are \emph{no} decision procedures for the validity of pure PCNF formulas whose
prefixes have one of the following forms:
\[ \exists z \forall x_1\cdots \forall x_n \exists y_1 \cdots \exists y_m, \]
\[ \forall x_1\cdots \forall x_n \exists y_1 \exists y_2 \exists y_2 \forall z_1 \cdots \forall z_m. \]
For the first prefix, the result holds even if $n = m = 1$:
\[ \exists z \forall x_1 \exists y_1, \]
and for the second prefix, the result holds even if $n = 0, m = 1$:
\[ \exists y_1 \exists y_2 \exists y_2 \forall z_1. \]
Even if the matrix is restricted to contain only binary predicate symbols, there
is still no decision procedure.
\end{thm}
\end{wideslide}

\begin{wideslide}[bm=,toc=]{Application to Policy Statements}
\begin{ex}{3.2}[Halpern and Weissman]
Consider the policy `anyone who is accompanied by a librarian
may enter the stacks'.\\~\\
In first-order logic:
\begin{align*}
& \forall x_1 (\exists x_2 (\bid{Librarian}(x_2) \land \bid{Accompanies}(x_2,x_1)) \imp \\
& \bid{Permitted}(x_1, enter(stacks)))
\end{align*}
\end{ex}
\begin{itemize} 
\item Note that \emph{enter} is a function.
\item Validity for existential formulas with functions is undecidable. 
\item Need a more restricted sublanguage.
\item Decidable is not enough---must also be \emph{tractable}.
\end{itemize}
\end{wideslide}

\begin{wideslide}[bm=,toc=]{A Tractable Sublanguage for Access
  Control}
Halpern and Weissman propose and analyze the complexity and expressiveness
of several fragments of FOL. One example:
\begin{thm}{4.2}[Halpern and Weissman]
Let $\Phi$ be a vocabulary that contains {\bf Permitted} (and possibly
other predicate, constant, and function symbols). Let $\mathcal{L}_6$
consist of all closed formulas in $\mathcal{L}^{fo}(\Phi)$ of the
form $E \land P \imp \bid{Permitted}(t,t')$, where $P$ is a conjunction
of standard policies and both $t$ and $t'$ and closed terms of the
appropriate sort, such that
\end{thm}
\vspace{-2ex}
\begin{enumerate}
\renewcommand{\labelenumi}{\alph{enumi})}
\item E is a basic environment with $m$ constants,
\item no policy in P has an inequality in its antecendent, and
\item there are no bipolars in P relative to the equality statements
in E.
\end{enumerate}
\emph{If each literal in each policy has at most one variable
  that does not appear in {\bf Permitted}, then we can determine
    the validity of the formula in time $O((|E| + m|P|)log|E|)$.}

    {\tiny (Refer to paper for detailed explanation of these constraints.)}
\end{wideslide}


\end{document}

