\documentclass[style=sailor,size=12pt]{powerdot}
\usepackage{epic,array,ecltree,url,calrsfs}
\usepackage[nointegrals]{wasysym}
\usepackage{listings}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsxtra}
\usepackage{amsthm}
\usepackage{mlextra} % Must be below ams packages
\usepackage{mathrsfs}
\usepackage{color}
\usepackage{array}
\usepackage{graphicx}
\graphicspath{ {../art/} }
\usepackage{bm}
\usepackage{tikz}
\usepackage{multicol}
\usepackage{enumitem}

\pdsetup{method=normal}

\title{First Order Logic: Part I}
\author{Foundations of Computer Science}
\date{\today}

\begin{document}
\maketitle
\begin{slide}[toc=,bm=]{Overview}
This slide set covers the following topics in first order logic:

\vspace{5mm}
\tableofcontents[content=sections]
\end{slide}

\section[slide=true]{Motivations and Applications}
\include{fol_why}

\section[slide=true]{Syntax}
\include{fol_syntax}

\section[slide=true]{Semantics}
\include{fol_semantics}

\section[slide=true]{Validity and Satisfiability}

\begin{wideslide}[bm=,toc=]{Validity and Satisfiability}
\begin{defn}{7.23}[Ben Ari]
Let $A$ be a closed formula of first-order logic.
\end{defn}
\vspace{-2ex}
\begin{itemize}
\item<2-> $A$ is \emph{true} in $\mathcal{I}$ or $\mathcal{I}$ is a \emph{model} for
$A$ iff $v_{\mathcal{I}}(A) = T$. 
\begin{itemize}
\item<3-> Notation: $\mathcal{I} \models A$.
\end{itemize}
\item<4-> $A$ is \emph{valid} if for \emph{all} interpretations $\mathcal{I}$,
$\mathcal{I} \models A$. 
\begin{itemize}
\item<5-> Notation: $\models A$ 
\end{itemize}
\item<6-> $A$ is \emph{satisfiable} if for \emph{some} interpretation $\mathcal{I}$,
$\mathcal{I} \models A$.
\item<7-> $A$ is \emph{unsatisfiable} if it is not satisfiable. 
\item<8-> $A$ is \emph{falsifiable} if it is not valid. 
\end{itemize}

\end{wideslide}

\begin{wideslide}[bm=,toc=]{Satisfiability and Validity in $\forall x \; p(a,x)$}
Let $A$ be the formula $\forall x \; p(a,x)$ from examples $7.17$ and $7.21$:
\begin{enumerate}
\item<2-> $\mathcal{I}_1 = (\N_0,\{\leq\},\{0\})$
\begin{itemize}
\item<3-> $T$: $A$ is satisfiable. 
\end{itemize}
\item<4-> $\mathcal{I}_2 = (\N_0,\{\leq\},\{1\})$
\begin{itemize}
\item<5-> $F$: $A$ is not valid. 
\end{itemize}
\item<6-> $\mathcal{I}_3 = (\Z,\{\leq\},\{0\})$
\begin{itemize}
\item<7-> $F$: Also shows $A$ is not valid. 
\end{itemize}
\item<8-> $\mathcal{I}_4 = (\mathcal{S},\{substr\},\{ \epsilon \})$
\begin{itemize}
\item<9-> $T$: Also shows $A$ is satisfiable. 
\end{itemize}
\end{enumerate}
\end{wideslide}


\begin{wideslide}[bm=,toc=]{Satisfiability and Validity Examples}
\begin{defn}{7.25}{Ben Ari}
\end{defn}
\begin{enumerate}
\item<2-> $\forall x \forall y (p(x,y) \imp p(y,x))$
\begin{itemize}
\item<3-> Satisfiable? \pause[3] Yes, in interpretations where $p$ is a
symmetric relation.
\item<3-> Valid? \pause No. Falisified in interpretations where $p$ is
not symmetric. 
\end{itemize}
\item<6-> $\forall x \exists y \; p(x,y)$
\begin{itemize}
\item<7-> Satisfiable? \pause[3] Satisfiable in interpretations where 
$p$ is a total function. For example, $(x,y) \in R$ iff $y = x + 1$
for $x,y \in \Z$.
\item<7-> Valid? \pause No. Falsified if the domain is changed to $\Z^-$.  
\end{itemize}
\item<10-> $\exists x \exists y(p(x) \land \ngg p (y))$
\begin{itemize}
\item<11-> Satisfiable? \pause[3] Yes. Assume for example that $D$ is $\N$ and
$p$ is the unary predicate corresponding to the property of being an
odd number.
\item<11-> Valid? \pause No. Consider a domain with only one element.
\end{itemize}
\end{enumerate}
\end{wideslide}

\begin{wideslide}[bm=,toc=]{Satisfiability and Validity Examples (continued)}
\begin{defn}{7.25}{Ben Ari}
(continued from previous slide)
\end{defn}
\begin{enumerate}
\setcounter{enumi}{3}
\item<2-> $\forall x \; p(a,x)$
\begin{itemize}
\item<3-> Expresses the existence of an element with special properties.
\item<3-> Satisfiable? \pause[3] Yes, when $D$ is $\N$ and $p$ is $\leq$.
\item<3-> Valid? \pause No. Falisified if we change $D$ to $\Z$ 
\end{itemize}
\item<6-> $\forall x (p(x) \land q(x)) \eqv (\forall x p(x) \land \forall x
    q(x))$
\begin{itemize}
\item<7-> Satisfiable? \pause[3] Yes. Ex: $D$ is $\N$, $p$ is the property of
not being negative and $q$ is the property of being divisible by $1$. 
\item<7-> Valid? \pause Yes. Proof uses theorem 7.22. 
\end{itemize}
\item<10-> $\forall x (p(x) \imp q(x)) \imp (\forall x p(x) \imp \forall x q(x))$
\begin{itemize}
\item<11-> Valid? \pause Yes. 
\item<11-> What about the converse: $(\forall x p(x) \imp \forall x q(x)) \imp \forall x (p(x) \imp q(x))$?
\end{itemize}
\end{enumerate}
\end{wideslide}

\begin{wideslide}[bm=,toc=]{Extending to Sets of Formulas}
As in PL, we can extend concepts of interpretation, satisfiability and other
properties to sets of formulas.
\pause
\begin{defn}{7.26}[Ben Ari]
Let $U = \{A_1,...\}$ be a set of formulas where 
\begin{itemize}
\item<3-> $\{p_1,...,p_m\}$ are all the predicates appearing in all $A_i \in U$ and
\item<4-> $\{a_1,...,a_k\}$ are all the constants appearing in all $A_i \in U$. 
\end{itemize}
\pause[3]
An \emph{interpretation} $\mathcal{I}_U$ for $U$ is a triple:
\[(D, \{R_1,...,R_m\},\{d_1,...,d_k\})\]
\pause
where 
\begin{itemize}
\item<6-> $D$ is a \emph{non-empty} set called the \emph{domain}, 
\item<7-> $R_i$ is an $n_i$-ary relation on $D$ that is assigned to the $n_i$-ary predicate $p_i$ and 
\item<8-> $d_i \in D$ is an element of $D$ that is assigned to the constant $a_i$.
\end{itemize}
\end{defn}

\end{wideslide}

\begin{wideslide}[bm=,toc=]{Consistency}
\begin{defn}{7.27}[Ben Ari]~\\~\\
A set of closed formulas $U = \{A_1,...\}$ is \emph{(simultaneously)
  satisfiable} iff 
\begin{itemize}
\item<2-> there exists an interpretation $\mathcal{I}_U$ such that $v_{\mathcal{I}_U}(A_i) = T$ for all $i$. 
\item<3-> The satisfying interpretation is a \emph{model} of $U$.
\item<4-> $U$ is \emph{valid} iff for every interpretation
  $\mathcal{I}_U$, $v_{\mathcal{I}_U}(A_i) = T$ for all $i$.
\end{itemize}
\end{defn}
\end{wideslide}

\section[slide=true]{Logical Equivalence}

\begin{wideslide}[bm=,toc=]{Definition of Logical Equivalence}
\begin{defn}{7.28}[Ben Ari]
\end{defn}
\vspace{-2ex}
\begin{itemize}
\item<2-> Let $U = \{A_1,A_2\}$ be a pair of closed formulas. 
\item<3-> $A_1$ is \emph{logically equivalent} to $A_2$ iff \pause[3] $v_{{\mathcal{I}_U}}(A_1) =
v_{{\mathcal{I}_U}}(A_2)$ for \emph{all} $\mathcal{I}_U$.
\begin{itemize}
\item<5-> Notation: $A_1 \equiv A_2$.
\end{itemize}
\item<6-> Let $A$ be a closed formula and $U$ a set of closed formulas.
\item<7-> $A$ is a \emph{logical consequence} of $U$ iff \pause[4] for all interpretations
$\mathcal{I}_{U \cup \{A\}}$, $v_{\mathcal{I}_{U \cup \{A\}}}(A_i) = T$ for
all $A_i \in U$ implies $v_{\mathcal{I}_{U \cup \{A\}}}(A) = T$. 
\begin{itemize}
\item<9-> Notation: $U \models A$.
\end{itemize}
\end{itemize}
\vspace{-5mm}
\pause[2]
\begin{thm}{7.29}
Let $A,B$ be closed formulas and $U = \{A_1,...,A_n\}$ be a set of
closed formulas. Then:
\end{thm}
\vspace{-5ex}
\begin{tabbing}
closed formulas. Then X \= \kill
\> $A \equiv B$ iff $\models A \eqv B$\\
\> $U \models A$ iff $\models (A_1 \land \cdots \land A_n) \imp A$
\end{tabbing}

\end{wideslide}

\begin{wideslide}[bm=,toc=]{Duality}
The two quantifiers are duals:
\pause
\begin{tabbing}
The two quantifiers are dua \= \kill
\> $\models \forall x A(x) \eqv \ngg \exists x \ngg A(x)$.\\
\> ~\\ 
\pause
\> $\models \exists x A(x) \eqv \ngg \forall x \ngg A(x)$.
\end{tabbing}
\pause
\begin{itemize}
\item It is only necessary to define one of the two.
\item If $\forall$ is defined, $\exists$ can be considered an abbreviation of
      $\ngg \forall \ngg$.
\end{itemize}

\end{wideslide}

\begin{wideslide}[bm=,toc=]{Commutativity}
Quantifiers of the same type commute:
\pause
\begin{tabbing}
Quantifiers of the same \= \kill
\> $\models \forall x \forall y A(x,y) \eqv \forall y \forall x A(x,y)$.\\
\pause
\> ~\\ 
\> $\models \exists x \exists y A(x,y) \eqv \exists y \exists x A(x,y)$.\\
\end{tabbing}

\pause
but $\forall$ and $\exists$ commute only in one direction:
\pause
\begin{tabbing}
Quantifiers of the same \= \kill
\> $\models \exists x \forall y A(x,y) \imp \forall y \exists x A(x,y)$.\\
\end{tabbing}

\end{wideslide}

\begin{wideslide}[bm=,toc=]{Distributivity}
\begin{tabbing}
Universal \= \kill
Universal quantifiers distribute over conjunction:  \> \\[2mm]
\pause
\>$\models \forall x(A(x) \land B(x)) \eqv \forall x A(x) \land \forall x B(x)$.\\
~\\
\pause
Existential quantifiers distribute over disjunction:\\[2mm]
\pause
\>$\models \exists x(A(x) \lor B(x)) \eqv \exists x A(x) \lor \exists x B(x)$.\\
~\\
\pause
Distributing universal over disjunction goes only one direction:\\[2mm]
\pause
\>$\models \forall x A(x) \lor \forall x B(x) \imp \forall x(A(x) \lor B(x))$.\\
~\\
\pause
Distributing existential over conjunction goes the other direction:\\[2mm]
\pause
\>$\models \exists x(A(x) \land B(x)) \imp \exists x A(x) \land \exists x B(x)$.\\
\end{tabbing}

\end{wideslide}

\begin{wideslide}[bm=,toc=]{Quantification Without the Free Variable in Its Scope}

Quantifiers over disjunction or conjunction are always distributive when
one subformula does not contain the quantified variable:

\vspace*{-5mm}
\begin{displaymath}
\def\arraystretch{1.5}
\begin{array}{ll}
 \models \exists xA(x) \vee B \eqv \exists x(A(x) \vee  B), &
 \models \forall xA(x) \vee B \eqv \forall x(A(x) \vee  B), \\
 \models B \vee \exists xA(x) \eqv \exists x(B \vee A(x)), &
 \models B \vee \forall xA(x) \eqv \forall x(B \vee A(x)), \\
 \models \exists xA(x) \wedge B \eqv \exists x(A(x) \wedge B), &
 \models \forall xA(x) \wedge B \eqv \forall x(A(x) \wedge B), \\
 \models B \wedge \exists xA(x) \eqv \exists x(B \wedge A(x)), &
 \models B \wedge \forall xA(x) \eqv \forall x(B \wedge A(x)).
\end{array}
\end{displaymath}
\pause
For implication, the following equivalences hold (note they are not symmetric):
\begin{displaymath}
\begin{array}{l}
 \models \forall x(A\imp B(x)) \eqv (A \imp \forall xB(x)) \\ 
 \models \forall x(A(x)\imp B) \eqv (\exists xA(x) \imp B) \\
\end{array}
\end{displaymath}

\end{wideslide}
\begin{wideslide}[bm=,toc=]{Quantification Over Implication}
The following formulas hold for distributing quantifiers over implication:
\vspace*{-2mm}
\begin{displaymath}
\def\arraystretch{1.5}
\begin{array}{l}
\models \forall x(A(x)\imp B(x))\imp (\forall xA(x)\imp\forall xB(x)) \\
\models \forall x(A(x)\imp B(x))\imp (\exists xA(x)\imp\exists xB(x)) \\
\models \forall x(A(x)\imp B(x))\imp (\forall xA(x)\imp\exists xB(x)) \\
\models \exists x(A(x)\imp B(x))\eqv (\forall xA(x)\imp\exists xB(x))  \\
\models (\exists xA(x)\imp\forall xB(x))\imp \forall x(A(x)\imp B(x))  \\
\end{array}
\end{displaymath}
\pause
Example to verify intuition:
\begin{itemize}
\item Let $A(x)$ mean ``a person, $x$, is rich.''
\item Let $B(x)$ mean ``a person, $x$, is happy.''
\item Check the formulas under this interpretation.
\end{itemize}
\end{wideslide}

\begin{wideslide}[bm=,toc=]{Quantification Over Equivalence}
Distribution of universal quantification works in one direction:
\begin{displaymath}
\begin{array}{l}
 \models \forall x(A(x)\eqv B(x))\imp (\forall xA(x) \eqv \forall xB(x)) \\
\end{array}
\end{displaymath}
\pause
The following formula holds for existential quantification:
\begin{displaymath}
\begin{array}{l}
 \models \forall x(A(x)\eqv B(x))\imp (\exists xA(x) \eqv \exists xB(x)) \\
\end{array}
\end{displaymath}

\end{wideslide}

\begin{wideslide}[bm=,toc=]{Example Derivation for Distributing Quantifiers Over
Implications / Biconditionals}
To derive an equivalent formula, translate $\imp$ or $\eqv$ into disjuctions and
conjunctions, then apply relevance substitutions.
\begin{displaymath}
\def\arraystretch{1.5}
\begin{array}{ll}
\exists x (A(x) \imp B(x))  & \equiv \exists x(\ngg A(x) \lor B(x)) \\
\pause &\equiv \exists x\ngg A(x) \lor \exists x B(x) \\
\pause  &\equiv \ngg \exists x\ngg A(x) \imp \exists x B(x) \\
\pause &\equiv \forall x A(x) \imp \exists x B(x) \\

\end{array}
\end{displaymath}

\end{wideslide}

\section[slide=true]{Functions}
\include{fol_functions}

\end{document}

