
\documentclass[style=simple,size=12pt]{powerdot}
\usepackage{epic,array,ecltree,wasysym,url}
\usepackage{mlextra}

\newcommand{\id}[1]{\mbox{\it #1\/}}
\newcommand{\rid}[1]{\mbox{\rm #1}}
\newcommand{\sid}[1]{\mbox{\sf #1}}
\newcommand{\bid}[1]{\mbox{\bf #1}}
\newcommand{\tinysz}[1]{\mbox{\tiny $#1$}}

\pdsetup{method=normal}

\begin{document}

\begin{wideslide}[bm=,toc=]{Sample RT derivation}

\begin{center}
\begin{tabular}{llllll}
 & & $\bid{B}$ & $\id{nil}\in\id{RTL}$ & & \\ \cline{4-4}
$\bid{B}$ & $\id{nil}\in\id{RTL}$ & $\bid{R2}$ & $\id{node}(\id{nil})\in\id{RT}$ & $\id{nil}\in\id{RTL}$ & $\bid{B}$ \\ 
\cline{2-2} \cline{4-5}
$\bid{R2}$ & $\id{node}(\id{nil})\in\id{RT}$ & $\bid{R1}$ & \multicolumn{2}{l}{
$\id{cons}(\id{node}(\id{nil}),\id{nil})\in\id{RTL}$} & \\
\cline{2-5}
$\bid{R1}$ & \multicolumn{4}{l}{$\id{cons}(\id{node}(\id{nil}),\id{cons}(\id{node}(\id{nil}),\id{nil})) \in \id{RTL}$} \\
\cline{2-5}
$\bid{R2}$ & \multicolumn{4}{l}{$\id{node}(\id{cons}(\id{node}(\id{nil}),\id{cons}(\id{node}(\id{nil}),\id{nil})))\in\id{RT}$}
\end{tabular}
\end{center}
\end{wideslide}
\end{document}
 
 & $e,s \vdash 1 \rightarrow 1,e,s$ & & $l_1\in\id{dom}(s)$ & \\ \cline{2-2}\cline{4-4}
$\id{sum}$ & $e,s \vdash \ast\!\ast\! x+1 \rightarrow 4,e,s$ & & $e,s\vdash \ast x \rightarrow l_2,e,s$ & $\id{deref}$ \\ 
 & $l_2\in\id{dom}(s)$ & & & \\ \cline{2-4}
 & \multicolumn{3}{c}{$e,s\vdash \ast\!\ast\! x := \ast\!\ast\! x+1
\rightarrow 4,e,s[l_2\mapsto 4]$} & $\id{assign1}$
\end{tabular}
\end{center}

\item Proofs {\em about\/} the Post system prove properties of the relation,
or in other words about {\em all\/} programs expressed in the simple language.
For instance, one can show that the Post system over
$(\id{Env}\times\id{S}\times E\times V\times \id{Env}\times S)$
is decidable.
\item The Post system is r.e. but undecidable.
\item A semi-decision procedure for the Post system normally
answers {\em queries\/} of the form $e,s\vdash C\rightarrow x,y,z$ where
$x$, $y$ and $z$ are Post-system variables, and $e$ and $s$ are not
variables but rather concrete instances of an environment 
and store respectively as in the example above.
\item The Post system is deterministic.
For any expression or command in the simple language, exactly one 
rule applies to it.
This facilitates finding proofs in the system since an expression or command
can be compiled into a deterministic representation
and ``executed'' in order to help speed up a semi-decision procedure.
Of course the compilation must preserve what is provable within
the system, and this is the goal of compiler verification.
\end{itemize}

\end{document}
